Dotted Suffix Trees\\A Structure for Approximate Text Indexing

### 2005-07-28 paper/main.tex
Add abstract & introduction Ignore generated files Fill in biblio

\documentclass[a4paper,10pt]{article}
\usepackage[boxruled,vlined,portugues,figure,linesnumbered,longend]{algorithm2e}

\newcommand*{\Assign}{\ensuremath\longleftarrow}
\newcommand{\putstring}[1]{\textsl{#1}}
\newtheorem{definition}{Definition}

@@

\begin{abstract}

\end{abstract}

\section{Introduction}

\subsection{Related Work}
\subsection{Structure of the Paper}

In section~\ref{sec:structure} we present the structure and the way to use it for searching. In section~\ref{sec:build-1} we show the way to build a structure to handle one error only and analize this version of the algorithm. Then, we show, in section~\ref{sec:build-k} how to generalize this procedure for any number of errors and how the previous analisys also applies to any number of errors.

@@

\subsection{Intuition}
Suffix Trees have become a well known structure since their introduction in 1973~\cite{weiner}. In Figure~\ref{fig:suffix-0} we show the suffix tree for the string \putstring{mississipi}.

We add to each internal node in the structure a new link, which we call the \emph{error link}.

@@

@@

\begin{definition}[$\Sigma^+$ tree]
$T$ is a $\Sigma^+$ tree if $T$ is a rooted tree with edge labels from $\Sigma^+$. For each $a \in \Sigma$ and every node $n$ in $T$, there is at most an edge leaving $n$ whose label is $aw$.

$\Sigma^+$ trees are also called a PATRICIA trees~\cite{pat}.

@@

@@

@@

\begin{definition}{Suffix Link}
A suffix link in a suffix tree is a link from the node $\overline{aw}$ to the node $\overline{w}$. This node has the label $a$.
\end{definition}

@@

\begin{definition}{Position Set}
Given a node $\overline{w}$ in a suffix tree, it's \emph{position set} is the set formed by taking it's occurence set and adding the lenght of $w$ to each element.
\end{definition}

@@

How much space does a k-error dotted tree take? Suffix Trees have a number of nodes proportional to the size of the string (which we will refer to as $n$). One-error dotted trees have more nodes. The worst case presented in Figure~\ref{fig:aaaa} is a string of the form \putstring{aa\ldots{}aa} which generates a one-dotted tree taking $O(n^2)$ nodes. For a larger number of errors $k$, it is easy to see that this leads to $O(n^{k+1})$ nodes.

@@

>>>>

\documentclass[letter,10pt]{article}
\usepackage[boxruled,vlined,portugues,figure,linesnumbered,longend]{algorithm2e}
\input{figures/transfig}
\usepackage[]{graphicx}

\newcommand*{\Assign}{\ensuremath\longleftarrow}
\newcommand{\putstring}[1]{\textsl{#1}}
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}

@@

\begin{abstract}
\input{abstract}
\end{abstract}

\section{Introduction}
\input{introduction}

@@

\subsection{Intuition}

\begin{figure}
\includegraphics{figures/mississipi-0}
\caption{Suffix Tree for the string \putstring{mississipi}}%
\label{fig:mississipi-0}
\end{figure}

Suffix Trees have become a well known structure since their introduction in 1973~\cite{weiner}. In Figure~\ref{fig:mississipi-0} we show the suffix tree for the string \putstring{mississipi}.

Being at a node in the suffix tree corresponds to being in several points in the string at the same time. We now add to each node an extra edge which corresponds to moving one character forward in all these points.

@@

We assume that the reader is familiar with the basic concepts of suffix trees and strings and present the definitions below mainly to introduce the particular notation and terminology used in this paper.

@@

\begin{definition}[Patricia tree]
$T$ is a Patricia tree if $T$ is a rooted tree with edge labels from $\Sigma^+$. For each $a \in \Sigma$ and every node $n$ in $T$, there is at most an edge leaving $n$ whose label is $aw$.

@@

The name Patricia is sometimes spelt PATRICIA, since it was formed from the initial of Practical Algorithm to Retrieve Information Coded in Alphanumeric~\cite{morrison:patricia}.

@@

In this paper, we will be interested in substitution of Hamming distance:

@@

\begin{definition}[Suffix Link]
A suffix link in a suffix tree is a link from the node $\overline{aw}$ to the node $\overline{w}$. This node has the label $a$.
\end{definition}

@@

\begin{definition}[Position Set]
Given a node $\overline{w}$ in a suffix tree, it's \emph{position set} is the set formed by taking it's occurence set and adding the lenght of $w$ to each element.
\end{definition}

@@

\begin{figure}
\includegraphics{figures/aaaa}
\caption{$a^n$: worst case space consumption}%
\label{fig:aaaa}
\end{figure}

How much space does a k-error dotted tree take? Suffix Trees have a number of nodes proportional to the size of the string (which we will refer to as $n$). One-error dotted trees have more nodes. The worst case presented in Figure~\ref{fig:aaaa} is a string of the form $a^n = \underbrace{aaa\cdots}_{n \mathit{times }a}$ which generates a one-dotted tree taking $O(n^2)$ nodes. For a larger number of errors $k$, it is easy to see that this leads to $O(n^{k+1})$ nodes.

\begin{definition}[weight under a node, weight of a tree]
The weight under a node $\mathit{Weight}(n)$ is the number of leafs in the subtree rooted at that node.

The total weight of a tree is the sum of the weights of all the nodes.
\end{definition}

\begin{definition}[height of a node, average height]
The height of a node $\mathit{Height}(n)$ is the number of nodes in the path leading to that node from the root. The average height of the tree $\hat{h}$, is the average height of its nodes.

\[ \hat{h} = \frac{1}{n} \sum_{x \mathit{node of \mathcal{T}}} \mathit{Height}(x), n \mathrm{being the number of nodes} \]
\end{definition}

\begin{lemma}
Given a tree (any tree, not just suffix or similiar tree), the sum of the weights of all the nodes in the tree is equal to the average height of the node times the number of nodes.
\end{lemma}

Consider the tree being built node by node, in any order, as long as the property that it remains a tree during the process is maintained. Each node added, adds a certain amount of weight to the tree. In particular it adds one for each node that sits above it in the tree. This is the same as saying that each node adds its height (which does not change). The weight of the final tree is the sum of all these values.

\[ \mathit{Weight}(\mathcal{T}) = \sum_{x \in \mathcal{T}}\mathit{Height}(x) = n\frac{1}{n}\sum_{x \in \mathcal{T}}\mathit{Height}(x) = n\hat{h} \]

FIXME

@@

\section{Experimental Results}

\section{Future Work, Open Problems}

In the example for the string \putstring{mississipi}, presented in Figure~\ref{fig:mississipi-1}, one can see that the tree below \putstring{s.i} and \putstring{ssi} are exactly the same. Whether such occurences are the basis for a significant space saving and what algorithms mightexploit them is an open question.
Along the same lines, the structure's definition might be extended to structures such as the suffix-DAG presented in~\cite[7.7]{gusfield:algorithms}. An algorithm to efficiently construct it over these suffix-DAGs is another open problem.
